---
title: "MP1"
author: "Emma Laslo, Jonah Kotzen, Marc Godec"
date: "2024-01-30"
output: html_document
---


```{r setup, include=FALSE}

# library set up

knitr::opts_chunk$set(echo = TRUE)
library(av)
library(tidyverse)
library(ggplot2)

```

```{r chunk 2, include=FALSE, eval=FALSE}

# DO NOT RUN!!!!!! (this is for movies to avi)

f <- list.files(full.names = T,pattern=".MOV")

#create a place to store images
dir.create("images")

#for each file, f, breakup the .MOV into the images and build avi files
for(i in f){
  if(dir.exists("images")) unlink("images",recursive = T)

  #extract images
  av_video_images(i,destdir ="images",format="tiff")
  
  #list images
  f.i <- list.files("images",full.names = T)
  
  #make the avi from the images from the MOV
  av_encode_video(f.i,gsub("MOV","avi",basename(i)),codec = "rawvideo")
}
```


# Reading in Data

```{r chunk 3, include=TRUE}
walk.files <- list.files(pattern="walk_")
print(walk.files)

walk.dat <- list() #make a list for storing
for(i in walk.files){
  walk.i <- read_delim(i,delim="\t") #read data as a tibble
  met.dat <- unlist(strsplit(i,"_")) #split i for metadata and unlist it
  speed <- met.dat[2] #get speed
  subject <- gsub(".txt","",met.dat[3]) #remove ".txt" 
  walk.dat[[i]] <- walk.i%>%
    mutate(speed=speed,subject=subject)
}

walk.dat <- do.call(rbind,walk.dat)
head(walk.dat)

colnames(walk.dat) <- str_remove(colnames(walk.dat)," \\[pixel\\]| \\[sec\\]")
head(walk.dat)

unique(walk.dat$subject)
```

# Graphing Data 


## Flipped
```{r chunk 3.5, include=TRUE}
walk.dat2 <- walk.dat %>%
  select(TID, PID, x, y, subject, speed) %>%
  filter(TID != 3) %>%
  group_by(speed, subject) %>%
  mutate(y = abs(y - max(y)))  # Note: Adjust this if you need a different transformation

walk.dat2 %>%
  ggplot(aes(x, y, color = speed)) +
  geom_point() +
  facet_grid(subject ~ .)
```

# Manipulating Data

## Correlating TIDs to Regions
```{r chunk 4, include=TRUE}
walk.dat2 <- walk.dat2%>%
  mutate(TID=replace(TID, TID==1, "hipR"))%>%
  mutate(TID=replace(TID, TID==2, "hipC"))%>%
  mutate(TID=replace(TID, TID==3, "legR"))
head(walk.dat2)
```

## Pivoting Wider
```{r chunk 5, include=TRUE}
walk.wide <- walk.dat2%>%
  pivot_wider(names_from=TID,values_from=c("x","y"))
head(walk.wide)
```

## Calculating Pelvic Tilt (Theta)
```{r chunk final, include=TRUE}
# Assuming walk.wide is our data frame after using pivot_wider()
walk.wide <- walk.dat2 %>%
  pivot_wider(names_from = TID, values_from = c("x", "y"))
# Calculating the angle theta
walk.wide <- walk.wide %>%
  mutate(
    dx = x_hipR - x_hipC,  # Change in x-coordinate
    dy = y_hipR - y_hipC,  # Change in y-coordinate
    theta = atan2(dy, dx) * (180 / pi)  # Calculate the angle in degrees
  )
# Viewing the first few rows of the data with theta
head(walk.wide)
```

## Calculating Mean Max Tilt
```{r chunk final final, include=TRUE}
# Calculate the maximum pelvic tilt angle for each subject and each speed
max_tilt_per_subject_speed <- walk.wide %>%
  group_by(subject, speed) %>%
  summarize(max_tilt = max(theta, na.rm = TRUE))

# Calculate the mean of the maximum tilt angles for each subject and each speed
mean_max_tilt_per_subject <- max_tilt_per_subject_speed %>%
  group_by(subject, speed) %>%
  summarize(mean_max_tilt = mean(max_tilt, na.rm = TRUE))

# Viewing the result
print(mean_max_tilt_per_subject)

```

## Graphing Max Tilt Between Speeds
```{r chunk tlt}

ggplot(mean_max_tilt_per_subject, aes(x = subject, y = mean_max_tilt, fill = speed)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Mean Maximum Pelvic Tilt Angle by Subject and Speed",
       x = "Subject",
       y = "Mean Maximum Pelvic Tilt Angle (degrees)",
       fill = "Speed") +
  theme_minimal()

```


# Areas to Address

## Reading in Class Data

```{r chunk class, include=TRUE} 

class_data <- read.csv('/Users/jonahkotzen/Desktop/MP1/class_gait_data.csv')

# Clean the 'sex' variable and remove NAs
class_data <- class_data %>%
  mutate(sex = as.factor(trimws(sex))) %>%
  filter(!is.na(sex))

# Check unique values again
unique(class_data$sex)


```

## Comparison of Theoretical and Actual Walking Speeds:


```{r walk speeds, include=TRUE}

# Compare vmax to recorded walking speeds
ggplot(class_data, aes(x = vmax.in.ms..1., y = recorded.walking.in.ms..1.)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Comparison of vmax and Recorded Walking Speeds",
       x = "Theoretical vmax (m/s)", y = "Recorded Walking Speed (m/s)")

```

## Patterns Revealed by Calculating the Froude Number:

```{r chunk froude, include=TRUE} 

# Plot Froude number against walking speed
ggplot(class_data, aes(x = Froude.number, y = recorded.walking.in.ms..1.),) +
  geom_point() +
  labs(title = "Froude Number vs Walking Speed",
       x = "Froude Number", y = "Walking Speed (m/s)")

```

## Variation of Actual Pendulum Walking Speed with Human Length and Sex

```{r chunk pend, include=TRUE} 

# Plot walking speed against leg length colored by sex with a best fit line
ggplot(class_data, aes(x = leg.length.in.m, y = recorded.walking.in.ms..1., color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Walking Speed vs Leg Length by Sex",
       x = "Leg Length (m)", y = "Walking Speed (m/s)")
```
## Variation of Actual Pendulum Walking Speed with Human Size and Sex

```{r chunk pends, include=TRUE} 

# Plot walking speed against leg length colored by sex with a best fit line
ggplot(class_data, aes(x = Mass.in.Kg, y = recorded.walking.in.ms..1., color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Walking Speed vs Mass by Sex",
       x = "Mass (kg)", y = "Walking Speed (m/s)")

```


## Changes in Pelvic Tilt with Increased Walking Speeds
```{r pelvtilt, include=TRUE}

ggplot(class_data, aes(x = normal.speed.mean.maximum.pelvic.tilt.angle, y = fast.speed.mean.maximum.pelvic.tilt.angle, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Pelvic Tilt Angle at Different Speeds by Sex",
       x = "Normal-Speed Mean Maximum Pelvic Tilt Angle (degrees)",
       y = "Fast-Speed Mean Maximum Pelvic Tilt Angle (degrees)")



```

## Variation of Pelvic Tilt at Normal Walking Speeds with Human Size and Sex:

```{r pelvtiltvar, include=TRUE}

# Create the first plot for normal walking speed vs. leg length
ggplot(class_data, aes(x = leg.length.in.m, y = normal.speed.mean.maximum.pelvic.tilt.angle, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Normal-Speed Pelvic Tilt vs Leg Length by Sex",
       x = "Leg Length (m)", y = "Normal-Speed Mean Maximum Pelvic Tilt Angle (degrees)") +
  theme(legend.position="none") # Hide legend to avoid duplication

# Create the second plot for normal walking speed vs. mass in Kg
ggplot(class_data, aes(x = Mass.in.Kg, y = normal.speed.mean.maximum.pelvic.tilt.angle, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Normal-Speed Pelvic Tilt vs Mass by Sex",
       x = "Mass in Kg", y = "Normal-Speed Mean Maximum Pelvic Tilt Angle (degrees)")

```


# Interpretations

